Autocorrelation in Sparse Distributed Memory

### 2019-07-18 Chapters/sdm-autocorrelation.tex
Papaer
wtf
quase draft

TRANSITION\footnote{Transition to what's coming next.}

@@

\caption{The threshold size $r$ of hard-locations bring about the autocorrelation of Theorem \ref{T1}. Consider, for instance, the extremes \{0, $n$\}.  Let us start with $n$ (in our case, $n=1000$ dimensions): given the information that $d(x,y)\le n$, the probability of a bitmatch in dimension $i$ is $1/2$; as $\forall x,y \in \{0, 1\}^N, d(x,y)\le n$. At the other extreme, consider that we have the information that $d(x,y)\le 0$: in this case $x=y$ and the probability of a bitmatch, in any dimension, is 1. The autocorrelation hence drops monotonically until convergence at $1/2$ as the distance grows. Numerically, our results converge to precisely $1/2$ only after $d(x,y)\ge 607$ for $n=1000$; $d(x,y)\ge 5332$ for $n=10000$, and $d(x,y)\ge 83$ for $n=100$. This will be seen in Lemma \ref{thm:sdm-access-radius}.
\label{fig:bitmatchbias}}
\end{figure}

\begin{lemma}
Let $r$ be the access radius given that $f$ percent of the hard locations are activated. Then, $\lim_{n \rightarrow \infty} r/n = 1/2$\footnote{AS WRITTEN, this cannot be right.  This is saying that, as the number of dimensions explodes, o raio sobre o número de dimensões converge p/ 1/2?  Não, converge para zero.  Temos que re-escrever.}.
\label{thm:sdm-access-radius}
\end{lemma}
\begin{proof}

@@

Let $\Phi(x)$ be the cdf of the standard normal distribution. Let $z = \frac{r - n/2}{\sqrt{n}/2}$. \footnote{$z$ is, hence, measuring the number of standard deviations from the average in the Gaussian, as $r$ varies.}  Thus, $P(d(x, y) \le r) = \Phi(z)$. As $f = P(d(x, y) \le r)$, then, $f = \Phi(z)$\footnote{WTF?  $f$ isn't a percentage here, is it?}.

@@

so what?\footnote{WHAT DOES THIS MEAN?}

what now?\footnote{AND WHAT HAPPENS NOW?}

@@

In our case, $\left[ \text{acc}_i|\eta_i=1 \right] \sim \mathcal{N}(\mu = 128.62, \sigma^2 = 12865.69)$, and $\left[ \text{acc}_i|\eta_i=0 \right] \sim \mathcal{N}(\mu = -128.62, \sigma^2 = 12865.69)$. See Figure \ref{fig:sdm-read} --- we can notice that the variance issue from Figure \ref{fig:sdm-read-sums} has propagated to these images.

@@

In order to check this probability, I have run a simulation reading from 1,000 random bitstrings (which have never been written into memory) and calculate the distance from the result of a single read. As the $P(wrong) = 0.22037$, I expected to get an average distance of 220.37 with a standard deviation of 13.10. See Figure \ref{fig:sdm-read-random-bs} for the comparison between the simulated and the theoretical outcomes.

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[width=0.5\textwidth]{./images02/autocorrelation/single-read-dist.png}

  \caption{New distance after a single read operation in a bitstring $\eta_d$, which is $d$ bits away from $\eta$. The new distance was calculated between $\eta_d$ and $\text{read}(\eta_d)$. Notice that when $d \ge 520$, the intersection between $\eta$ and $\eta_d$ is zero, which means there is only random bitstrings written into the activated hard locations. The distance 220 equals $1000 \cdot 0.220$ which is the probability find in Figure \ref{fig:sdm-read-random-bs}.}
  \label{fig:sdm-single-read-dist}
\end{figure}

>>>>

@@

\caption{The threshold size $r$ of hard-locations bring about the autocorrelation of Theorem \ref{T1}. Consider, for instance, the extremes \{0, $n$\}.  Let us start with $n$ (in our case, $n=1000$ dimensions): given the information that $d(x,y)\le n$, the probability of a bitmatch in dimension $i$ is $p=1/2$; as $\forall x,y \in \{0, 1\}^N, d(x,y)\le n$. At the other extreme, consider that we have the information that $d(x,y)\le 0$: in this case $x=y$ and the probability of a bitmatch, in any dimension, is $p=1$. The autocorrelation hence drops monotonically until convergence at $p=1/2$ as the distance grows.
\label{fig:bitmatchbias}}
\end{figure}

A natural question is what happens to this pull bias as $n$ grows?  Let $f = P(d(x,y)\le r) = 2^{-n}\sum_{k=0}^{r}{\binom{n}{k}}$.  Then, we have:

\begin{lemma}
Let $r$ be the access radius given that $f$ is the ratio of the hard locations activated. Then, $\lim_{n \rightarrow \infty} r/n = 1/2$.
\label{thm:sdm-access-radius}
\end{lemma}
\begin{proof}

@@

Let $\Phi(x)$ be the cdf of the standard normal distribution. Let $z = \frac{r - n/2}{\sqrt{n}/2}$. $z$ is, hence, measuring the number of standard deviations from the average in the Gaussian, as $r$ varies\footnote{MARCELO, CHEQUE ESTA ULTIMA FRASE PLEASE?}.  Thus, $P(d(x, y) \le r) = \Phi(z)$. As $f = P(d(x, y) \le r)$, then, $f = \Phi(z)$ for our fixed $r$}.

@@

As seen on Fig. \ref{fig:bitmatchbias}, the bias disappears as $r$ grows.  Numerically, our results converge to precisely $p=1/2$ only after $d(x,y)\ge 83$ for $n=100$, $d(x,y)\ge 607$ for $n=1000$; $d(x,y)\ge 5332$ for $n=10000$. This can be seen in Lemma \ref{thm:sdm-access-radius}.

@@

In our case, $\left[ \text{acc}_i|\eta_i=1 \right] \sim \mathcal{N}(\mu = 128.62, \sigma^2 = 12865.69)$, and $\left[ \text{acc}_i|\eta_i=0 \right] \sim \mathcal{N}(\mu = -128.62, \sigma^2 = 12865.69)$. See Figure \ref{fig:sdm-read} --- we can notice that the variance issue from Figure \ref{fig:sdm-read-sums} propagates directly here.

@@

In order to check this probability, we have run a simulation reading from 1,000 random bitstrings (which have never been written into memory) and calculated the distance from the result of a single read. As the $P(wrong) = 0.22037$, we expected to get an average distance of 220.37 with a standard deviation of 13.10. See Figure \ref{fig:sdm-read-random-bs} for the comparison between the simulated and the theoretical outcomes.

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[width=0.5\textwidth]{./images02/autocorrelation/single-read-dist.png}
  \caption{New distance after a single read operation in a bitstring $\eta_d$, which is $d$ bits away from $\eta$. The new distance was calculated between $\eta_d$ and $\text{read}(\eta_d)$. Notice that when $d \ge 520$, the intersection between $\eta$ and $\eta_d$ is zero, which means there is only random bitstrings written into the activated hard locations. The distance 220 equals $1000 \cdot 0.220$ which is the probability found in Figure \ref{fig:sdm-read-random-bs}.}
  \label{fig:sdm-single-read-dist}
\end{figure}

### 2019-07-18 paperIEEE.tex
quase draft

\caption{Activated addresses inside access \protect \\
radius $r$ around the center address.\label{fig-addresses-inside-access-radius}}
\end{figure}

@@

Since a cue $\eta_{x}$ near the target bitstring $\eta$ shares many hard locations with $\eta$, SDM can retrieve data from imprecise cues. Despite this feature, it is very important to know how imprecise this cue could be while still giving accurate results. What is the maximum distance from our cue to the original data that still retrieves the right answer? An interesting approach is to perform a read operation with a cue $\eta_{x}$, that is $x$ bits away from the target $\eta$.  Then measure the distance from the read output and $\eta$. If this distance is smaller than $x$ we are converging. Convergence is simple to handle, just read again and again, until it converges to the target $\eta$. If this distance is greater than $x$ we are diverging.   Finally, if this distance equals $x$ we are in a tip-of-the-tongue process.  A tip-of-the-tongue psychologically happens when you know that you know, but you can't say what exactly it is. In SDM mathematical model, a tip-of-the-tongue process takes infinite time to converge. \citet{Kanerva1988} called this $x$ distance, where the read's output averages $x$, the critical distance. Intuitively, it is the distance from which smaller distances converge and greater distances diverge. In Figure \ref{fig-p1-p2-iterative-read}, the circle has radius equal to the critical distance and every $\eta_{x}$ inside the circle should converge.  The figure also shows convergence in four readings.

@@

@@

Our conclusion is that this work is of special relevance as it brings about some \emph{systemwide properties} that models need to take into account.  The first decision choice involved in building a system with SDM regards the size of $n$, the number of dimensions of the memory.  Here, complementing the work of XXXXXXXX\footnote{check this}, we show that as $n$ grows, autocorrelation tends to $0$.  But recall that as $n$ grows, the size of the critical distance also tends to $0$.  It is therefore crucial to select a value for $n$ in which the model will retain its convergence properties, with the understanding of the possible effects brought about by autocorrelation.   Systemwide properties such as these affect the entire behavior of the model and, as such, should always be taken into account by theoreticians and modelers alike.

@@

@@

\begin{IEEEbiography}[
  {
  \includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{./Chapters/Linhares.jpg}
  }
  ]
{Alexandre Linhares} received the Ph.D. degree from the National Institute of Space Research, 2001. He is an Associate Professor at the Center of Behavioral Research, EBAPE / Getulio Vargas Foundation.  He has published ~30 papers in journals such as \emph{Artificial Intelligence, Applied Intelligence, Biological Cybernetics, Cognitive Science, Cognitive Systems Research, Computers & Operations Research, IEEE Transactions on Evolutionary Computation, Information Sciences, Frontiers in Human Neuroscience,} among others.
\end{IEEEbiography}

>>>>

\caption{A planar projection of the space illustrates some of the mechanisms.  Activated addresses inside access radius $r$ around the center address.\label{fig-addresses-inside-access-radius}}
\end{figure}

@@

Since a cue $\eta_{x}$ near the target bitstring $\eta$ shares many hard locations with $\eta$, SDM can retrieve data from imprecise cues. Despite this feature, it is very important to know how imprecise this cue could be while still giving accurate results. What is the maximum distance from our cue to the original data that still retrieves the right answer? An interesting approach is to perform a read operation with a cue $\eta_{x}$, that is $x$ bits away from the target $\eta$.  Then measure the distance from the read output and $\eta$. If this distance is smaller than $x$ we are converging. Convergence is simple to handle, just read again and again, until it converges to the target $\eta$. If this distance is greater than $x$ we are diverging.   Finally, if this distance equals $x$ we are in a tip-of-the-tongue process.  A tip-of-the-tongue psychologically happens when you know that you know, but you can't say what exactly it is. In SDM's mathematical model, a tip-of-the-tongue process takes infinite time to converge. \citet{Kanerva1988} called the distance in which the read's output averages $x$, the critical distance. Intuitively, it is the distance from which smaller distances converge and greater distances diverge. In Figure \ref{fig-p1-p2-iterative-read}, the circle has radius equal to the critical distance and every $\eta_{x}$ inside the circle should converge.  The figure also shows convergence in four readings.

@@

It was previously believed that reading at a distance $d=n/2$ would retrieve items at the equator distance, but this is not necessarily true. We have shown the subtle, but previously unnoticed, effect of autocorrelation in sparse distributed memory.  We have also derived excellent analytical approximations of the system's behavior. In this concluding section, some comments i) on the implications of this work, on ii) Future research, and iii) on the availability of the code are offered.

@@

Our conclusion is that this work is of special relevance as it brings about some \emph{systemwide properties} that models need to take into account.  The first decision choice involved in building a system with SDM regards the size of $n$, the number of dimensions of the memory.  Here, complementing the work of \cite{sjodin_improving_1995, sjodin_convergence_1998} in Jaeckel's SDM design\cite{jaeckel1989class}, we show that as $n$ grows, autocorrelation tends to $0$.  But recall that as $n$ grows, the size of the critical distance also tends to $0$.  It is therefore crucial to select a value for $n$ in which the model will retain its convergence properties, with the understanding of the possible effects brought about by autocorrelation.   Systemwide properties such as these affect the entire behavior of the model and, as such, should always be taken into account by theoreticians and modelers alike.

@@

\begin{table}[ht]
\begin{tabular}{|p{3.5cm}|p{4cm}|}  %{|c|c|}
\hline
System 1	& System 2\tabularnewline
\hline
\hline
Unconscious Reasoning	& Conscious Reasoning\tabularnewline
\hline
Implicit	& Explicit\tabularnewline
\hline
Automatic	& Controlled\tabularnewline
\hline
Low Effort	& High Effort\tabularnewline
\hline
Large Capacity	& Small Capacity\tabularnewline
\hline
Rapid	& Slow\tabularnewline
\hline
Default Process	& Inhibitory\tabularnewline
\hline
Associative	& Rule-Based\tabularnewline
\hline
Contextualized	& Abstract\tabularnewline
\hline
Domain Specific	& Domain General\tabularnewline
\hline
Evolutionarily Old	& Evolutionarily Recent\tabularnewline
\hline
Nonverbal	& Linked to language\tabularnewline
\hline
Includes recognition, perception, orientation	& Includes rule following, comparisons, weighing of options\tabularnewline
\hline
Modular Cognition	& Fluid Intelligence\tabularnewline
\hline
Independent of working memory	& Limited by working memory capacity\tabularnewline
\hline
Non-Logical	& Logical\tabularnewline
\hline
Parallel	& Serial\tabularnewline
\hline
\end{tabular}

\caption{Dual process theories.}

\label{table:twosys}

\end{table}

\begin{figure}[ht!]
  \centering

  \subfloat[Bongard problem 71]{\includegraphics[width=0.35\textwidth]{./images02/Bongard-problemp071.png}}

  \subfloat[Bongard problem 72]{\includegraphics[width=0.35\textwidth]{./images02/Bongard-problemp072.png}}

  \subfloat[Bongard problem 73]{\includegraphics[width=0.35\textwidth]{./images02/Bongard-problemp073.png}}

  \subfloat[Bongard problem 74]{\includegraphics[width=0.35\textwidth]{./images02/Bongard-problemp074.png}}

%  \subfloat[Bongard problem 75]{\includegraphics[width=0.35\textwidth]{./images02/Bongard-problemp075.png}}

%  \subfloat[Bongard problem 76]{\includegraphics[width=0.35\textwidth]{./images02/Bongard-problemp076.png}}

  \caption{Bongard problems 71 - 74, courtesy of Dr. Harry Foundalis\cite{foundalis_phaeaco:_2006, douglas1980godel, bongard_recognition_1968, Linhares2000}.  What distinguishes the boxes on the left hand side to those on the right hand side?  This is the quintessential `system-2' type of problem that current lies beyond AI's capabilities. The reader should note that here, for each class, $n=6$, a hard task for Machine Learning.}

  \label{fig:bongard-problems-74}
\end{figure}

@@

\begin{IEEEbiography}[
  {
  \includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{./Chapters/Linhares.jpg}
  }
  ]{Alexandre Linhares} received the Ph.D. degree from the National Institute of Space Research, 2001. He is an Associate Professor at the Center of Behavioral Research, EBAPE / Getulio Vargas Foundation.  He has published ~30 papers in journals such as \emph{Artificial Intelligence, Applied Intelligence, Biological Cybernetics, Cognitive Science, Cognitive Systems Research, Computers & Operations Research, IEEE Transactions on Evolutionary Computation, Information Sciences, Frontiers in Human Neuroscience,} among others.
\end{IEEEbiography}
